Evolution of Processes in Delivering Reliable Software within the European Scientific Arena

### 2018-08-31 paper-ieee-trans.tex
added a possible answer to editor s comments for the lack of quantification

\begin{document}
\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities, including physics, biology and
earth sciences, with intensive computing and large--scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

Without previous experience in such widely
collaborative projects, a big effort was spent in devising solutions for several
challenges \cite{datagrid}, namely 1) the communication overhead, as a
result of the large geographical separation of the involved parties in the
development tasks, 2) the fast evolution of the requirements from the user
communities (50 use cases from the three scientific area), and 3) the lack of a body
of knowledge for academic software engineering.

@@

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
shows a decrease in the unit testing coverage values for several products. As unit tests cover low-level
code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
We have observed that this decrease was aligned with high-demanding periods of software development, as
seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
density that this fact brought along, the overall performance kept progressing as the rest of the software
stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
The individual records followed also a growing trend (only 6 out of 32 components had lower values in 
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
implemented from scratch within the project. Software tools and libraries from 
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

The practical implementation of the CI scenario leverage from a tightly integrated open
source tools such as GitHub \cite{foot9}, as the online source code repository,

Jenkins is an event--response CI and Docker container
provisioning system that provides instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update.

@@

>>>>

\begin{document}
\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena}

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities (including physics, biology and
earth sciences) with intensive computing and large--scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

Without previous experience in such widely
collaborative projects, a big effort was spent in devising solutions for several
challenges \cite{datagrid}, namely 1) the communication overhead (as a
result of the large geographical separation of the involved parties in the
development tasks), 2) the fast evolution of the requirements from the user
communities (50 use cases from the three scientific area), and 3) the lack of a body
of knowledge for academic software engineering.

@@

On behalf of the quality assurance group, specific measures for the software development process and quality assurance were defined and introduced in the project. Through project monitoring and reporting mechanisms there was a successful control of the risks, efforts and document produced. The definition of a project developer's guide \cite{datagridguide} was able to provide developers with guidelines for the various software development phases (such as packaging, test and validation process, and style and naming convention) and to allow quality group to control software release via a procedure checklist.
The specification of quality indicators \cite{datagridindicators} allowed project managers to have an update view of the quality of the services over months. Figure \ref{fig:efficiency} shows the efficiency performance indicator measured as the number of jobs successfully completed divided the total number of jobs submitted: generally the project managers observed a positive trend of efficiency over months with only one exception for the LCG1 experiment due to some errors on the dedicated resources. Figure \ref{fig:bugzilla} shows the Bugzilla anomalies follow up in the second release.
\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/dgeff.png}
\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted): results are provided for five experiments involved in the projects.}
\label{fig:efficiency}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/dgbug.png}
\caption{Bugzilla anomalies follow up: new and pending anomalies are reported.}
\label{fig:bugzilla}
\end{figure}

@@

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the definition and periodic execution
of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's
internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
component between the two major releases, denoting an incremental trend over the course of the project:
code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
shows a decrease in the unit testing coverage values for 18\% of products. As unit tests cover low-level
code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
We have observed that this decrease was aligned with high-demanding periods of software development, as
seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
density that this fact brought along, the overall performance kept progressing as the rest of the software
stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
The individual records followed also a growing trend (only 6 out of 32 components had lower values in
INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments
implemented from scratch within the project. Software tools and libraries from
external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
analysis. However the values given apply to the entire codebase, even for the latter case of products. These
products challenged the application of the SQA policies described in this section, by two means:
\begin{itemize}
    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
\end{itemize}

@@

The practical implementation of the CI scenario leverage from tightly integrated open
source tools such as:
\begin{itemize}
    \item GitHub \cite{foot9} as the online source code repository hosting service,
    \item Jenkins as the event--response CI,
    \item Docker container provisioning system, to provide instant computing power needed
          for executing the pipeline jobs.
\end{itemize}
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update.

@@

\bibitem{datagridguide}
Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
[Accessed 30 Aug. 2018].

\bibitem{datagridindicators}
DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
2003.
[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
[Accessed 30 Aug. 2018].
